Pushdown Control-Flow Analysis via Refunctionalization (Functional Pearl)

### 2018-03-16 paper.tex
update

But, since we have added non-determinism into state, an empty list of frames means
we have the reached the halt of \textit{one computation path}, and we should
determine the next state indicates by @NDCont@.
So we have a pattern matching on @ndk@: 1) If the closure set is an empty set,
then we had tried all the closures of this fork point and should move the the
next fork point. Since the @NDCont@ is represented by a list and we always
append new element from its front, we are resuming to the most recent fork point by
popping up the front-most element.
2) Otherwise, we pop a closure from the set, then build new
environment and new store for the body expression of the closure;
we use the same frame list and time that are copied from the fork point rather
than current one.
The non-determinism continuation @new_ndk@ is also updated by removal of that closure.

@@

The shape for function @ndcontinue@ is similar to function @continue@, after
all, they both use @List@ as a representation of continuation.

@@

As we pointed earlier, refunctionalizing evaluation contexts to higher-order
continuation functions forces us to somehow sequentialize the abstract evaluation.
We identify that in order to sequentialize abstract evaluation, there are three
fundamental differences with concrete evaluation: 
1) The dereferencing store addresses is non-deterministic, therefore the returning
may also be non-deterministic.
2) We need sound collecting semantics rather than valuation semantics.
3) The abstract evaluation has to be terminating.

allocating continuations of defined language on the heap of defining language.

@@

Transform to Extended Continuation-Passing Style (ECPS), continuation and meta-continuation.
simply trun evaluation contexts into higher-order continuations.
We establish pushdown control-flow, but this is not computable.

\begin{lstlisting}
def nd[T,S](ts: Set[T], acc: S, f: (T, S, S => S) => S, g: S => S): S = {
  if (ts.isEmpty) g(acc)
  else f(ts.head, acc, (vss: S) => nd(ts.tail, vss, f, g))
}

type Ans = Set[VS]
type Cont = Ans => Ans

def aval(e: Expr, env: Env, store: BStore, time: Time, cont: Cont): Ans = {
  val new_time = (e::time).take(k)
  e match {
    case Let(x, App(f, ae), e) =>
      val closures = atomicEval(f, env, store).asInstanceOf[Set[Clos]]
      nd[Clos, Ans](closures, Set[VS](), { case (clos, acc, closnd) =>
        val Clos(Lam(v, body), c_env) = clos
        val baddr = allocBind(v, new_time)
        val new_env = c_env + (v -> baddr)
        val new_store = store.update(baddr, atomicEval(ae, env, store)) 
        aval(body, new_env, new_store, new_time, (bodyvss: Set[VS]) => {
          nd[VS, Ans](bodyvss, Set[VS](), { case (vs, acc_vss, bdnd) => 
            val VS(vals, time, store) = vs
            val new_env = env + (x -> baddr)
            val new_store = store.update(baddr, vals)
            aval(e, new_env, new_store, time, (evss: Ans) => bdnd(acc_vss ++ evss))
          },
          (evss: Ans) => closnd(evss ++ acc))
        })
      }, 
      cont)
    case ae if isAtomic(ae) => 
      val ds = atomicEval(ae, env, store)
      cont(Set(VS(ds, new_time, store))) 
  } 
}

def mtTime = List()
def mtEnv = Map[String, BAddr]()
def mtStore = Store[BAddr, Storable](Map())

def analyze(e: Expr) =
  aval(e, mtEnv, mtStore, mtTime, (ans => ans))
\end{lstlisting}

@@

\begin{lstlisting}
def aval(e: Expr, env: Env, store: BStore, time: Time, cache: Cache, cont: Cont): Ans = {
  val config = Config(e, env, store, time)
  if (cache.outContains(config)) return cont(Ans(cache.outGet(config), cache))

@@

  e match {
    case Let(x, App(f, ae), e) =>
      val closures = atomicEval(f, env, store).asInstanceOf[Set[Clos]]
      nd[Clos, Ans](closures, Ans(Set[VS](), new_cache), { case (clos, Ans(acc, cache), closnd) =>
        val Clos(Lam(v, body), c_env) = clos
        val baddr = allocBind(v, new_time)
        val new_env = c_env + (v -> baddr)
        val new_store = store.update(baddr, atomicEval(ae, env, store))
        aval(body, new_env, new_store, new_time, cache, { case Ans(bodyvss, bodycache) =>
          nd[VS, Ans](bodyvss, Ans(Set[VS](), bodycache), { case (vs, Ans(acc_vss, cache), bdnd) =>
            val VS(vals, time, store) = vs
            val baddr = allocBind(x, time)
            val new_env = env + (x -> baddr)
            val new_store = store.update(baddr, vals)
            aval(e, new_env, new_store, time, cache, { case Ans(evss, ecache) => bdnd(Ans(acc_vss ++ evss, ecache)) })
          },
          { case Ans(evss, cache) => closnd(Ans(evss ++ acc, cache)) })
        })
      },
      { case Ans(vss, cache) => cont(Ans(vss, cache.outUpdate(config, vss))) })

    case ae if isAtomic(ae) =>
      val vs = Set(VS(atomicEval(ae, env, store), new_time, store))
      cont(Ans(vs, new_cache.outUpdate(config, vs)))
  }
}
def analyze(e: Expr, env: Env = mtEnv, store: BStore = mtStore) = {
  def iter(cache: Cache): Ans = {
    val Ans(vss, new_cache) = aval(e, env, store, mtTime, cache, { 
      val initConfig = Config(e, env, store, mtTime)
      case Ans(vss, cache) => Ans(vss, cache.outUpdate(initConfig, vss)) 
    })
    if (new_cache.out == cache.out) { Ans(vss, new_cache) }
    else { iter(Cache(new_cache.out, new_cache.out)) }
  }    
  iter(mtCache)
}
\end{lstlisting}

@@

\begin{lstlisting}
def aval(e: Expr, env: Env, store: BStore, time: Time, cache: Cache): Ans = {
  val config = Config(e, env, store, time)
  if (cache.outContains(config)) {
    return Ans(cache.outGet(config), cache)
  }

  var new_cache = cache.outUpdate(config, cache.inGet(config))
  val new_time = (e::time).take(k)
  e match {
    case Let(x, App(f, ae), e) =>
      val closures = aeval(f, env, store).toList.asInstanceOf[List[Clos]]
      val letvs = for (Clos(Lam(v, body), c_env) <- closures) yield {
        val baddr = allocBind(v, new_time)
        val new_env = c_env + (v -> baddr)
        val new_store = store.update(baddr, aeval(ae, env, store))
        val Ans(bdv, bdcache) = aval(body, new_env, new_store, new_time, new_cache)
        new_cache = bdcache
        val evs = for (VS(vals, time, store) <- bdv) yield {
          val baddr = allocBind(x, time)
          val new_env = env + (x -> baddr)
          val new_store = store.update(baddr, vals)
          val Ans(ev, ecache) = aval(e, new_env, new_store, time, new_cache)
          new_cache = ecache
          ev
        }
        evs.foldLeft(Set[VS]())(_ ++ _)
      }
      val letv = letvs.foldLeft(Set[VS]())(_ ++ _)
      Ans(letv, new_cache.outUpdate(config, letv))

>>>>

But, since we have added non-determinism into state, an empty list of frames means
we have the reached the halt of \textit{one computation path}, and we should
determine the next state indicates by @NDCont@.
So we have a pattern matching on @ndk@:
\begin{itemize}
  \item If the closure set is an empty set,
then we had tried all the closures of this fork point and should move the the
next fork point. Since the @NDCont@ is represented by a list and we always
append new element from its front, we are resuming to the most recent fork point by
popping up the front-most element.
\item Otherwise, we pop a closure from the set, then build new
environment and new store for the body expression of the closure;
we use the same frame list and time that are copied from the fork point rather
than current one.
The non-determinism continuation @new_ndk@ is also updated by removal of that closure.
\end{itemize}

@@

The shape for function @ndcontinue@ is similar to function @continue@, after
all, they both use @List@ as a representation of continuations.
When the closure set is empty, we recursively call @drive_step@ with the rest
of @ndk@, since in this case the normal stack @konts@ is still empty, so
the state will be dispatched to @ndcontinue@ again.
Otherwise, we invoke @drive_step@ with the body expression of the next closure object.

@@

Refunctionalization transforms first-order data type represented
evaluation contexts to higher-order functions.
Besides sequentializing the order of abstract evaluation by functions,
this transformation can be also regarded as allocating continuation functions of
the defined language in the heap of the meta language.

In this section, several other notable changes have been made:
\begin{itemize}
\item We do not have the concept of @state@ (or @NDState@) anymore.
  The components of the states are lifted as arguments of the evaluation
  function.
\item Since our final target is abstract definitional interpreters while still
non-deterministic, start from this section, our abstract semantic artifact
returns a set of final values insteaf of collected states.
\item As we are getting close to the
abstract definitional interperters, the name of @drive_step@ is updated to
@aeval@ (for \emph{abstract eval}).
\end{itemize}

For clearity, in this section, we first show how to refunctionalize the normal continuation and
non-deterministic continuation without collecting semantics.
Then adding an appropriate caching algorithm to make sure the analysis will
terminate at some fixed-point is just changing the program to cache-passing style.
Finally, we examine how computable pushdown control-flow analysis is established
through refunctionalization and caching.

@@

To properly represented final values, we introduce the case class @VS@ which
contains a set of storable values, a time stamp, and a store.
\begin{lstlisting}
case class VS(vals: Set[Storable], time: Time, store: BStore)
\end{lstlisting}
An instance of @VS@ represents the computational result of one path in
the non-deterministic evaluation.
The reason that we include a store is there might be a case two pathes have the
same values but different accumulated side effects such as memory allocations.
Since the whole computation is non-deterministic, the type of final result
values is a collection @Set[VS]@. For convenience, we define it as answer type @Ans@:
\begin{lstlisting}
type Ans = Set[VS]
\end{lstlisting}

As we mentioned previously, the components of states are lifted as arguments to
@aeval@:
\begin{lstlisting}
def aeval(e: Expr, env: Env, store: BStore, time: Time, continue: Cont): Ans = ......
type Cont = Ans => Ans
\end{lstlisting}

An additional argument @continue@ which has type @Cont@ is also introduced.
Recall that in the disentangled AAM, the second case in function @drive_step@
makes a call to @continue@. Here, the reemerging @continue@ is a function
with type @Ans => Ans@. The function @aeval@ does not return, instead, it calls
@continue@ with the result values.

\textbf{Non-Determinism Abstraction.}
To handle the inevitable non-determinism, we defined a generic function @nd@ as follows:
\begin{lstlisting}
def nd[T,S](ts: Set[T], acc: S, f: (T, S, S => S) => S, g: S => S): S = {
  if (ts.isEmpty) g(acc)
  else f(ts.head, acc, (vss: S) => nd(ts.tail, vss, f, g))
}
\end{lstlisting}

Given a set of element of type @T@, an initial accumulated value of type @S@, a function @f@
that works on element type @T@ and accumulated type @S@, and a function @g@ that
works on and returns value of type @S@, @nd@ applies @f@ to each element in the
set iteratively, accumulates the value and finally applies @g@ on it. 
Seeing this one may bring @fold@ to mind, indeed, it is an implementation of
@fold@ in continuation style. But here we have two continuations, one for the
next element in the set, and one for the final result.
Sometime it is also called extended continuation-passing style (ECPS) that has a continuation and a
meta-continuation \cite{Danvy:1990:AC:91556.91622}.
\todo{correspondence to disentangled version, refunc the NDCont}

\textbf{First Version of Evaluation Function.}
Since our first version of @aeval@ just implements the non-deterministic
evaluation, and given the defined @nd@, the evaluation function can be easily
defined.
The outer call to function @nd@ evaluates over a set of possible closures of
@f@. For each closures, we go into its body with new environment and store.
Remember that function @aeval@ returns a set of @VS@ to its continuation,
so the @bodyvss@ is a set of @VS@ that represents values from possible
computation pathes when evaluating @body@.

Then for the inner call of function @nd@ evaluates over the set of body values.
For each @VS@, we use the time and store
The inner application of @aeval@ returns values of one computation path to its
continuation, where the continuation will accumulate the result and transfer
evaluation to the next body value.

\begin{lstlisting}
def aeval(e: Expr, env: Env, store: BStore, time: Time, continue: Cont): Ans = {
  val new_time = (e::time).take(k)
  e match {
    case Let(x, App(f, ae), e) =>
      val closures = atomicEval(f, env, store).asInstanceOf[Set[Clos]]
      nd[Clos, Ans](closures, Set[VS](), { case (clos, accouter, closnd) =>
        val Clos(Lam(v, body), c_env) = clos
        val baddr = allocBind(v, new_time)
        val new_env = c_env + (v -> baddr)
        val new_store = store.update(baddr, atomicEval(ae, env, store)) 
        aeval(body, new_env, new_store, new_time, (bodyvss: Set[VS]) => {
          nd[VS, Ans](bodyvss, Set[VS](), { case (vs, accinner, bdnd) => 
            val VS(vals, time, store) = vs
            val new_env = env + (x -> baddr)
            val new_store = store.update(baddr, vals)
            aeval(e, new_env, new_store, time, 
                  /* accumulate the values of one path, and call next body value*/
                  (evss: Ans) => bdnd(accinner ++ evss)) 
          },
          /* accumulate the values of pathes to accouter, and call next closure */
          (evss: Ans) => closnd(evss ++ accouter)) 
        })
      }, 
      continue)
    case ae if isAtomic(ae) => 
      val ds = atomicEval(ae, env, store)
      continue(Set(VS(ds, new_time, store))) 
  } 
}
\end{lstlisting}

At the end, the initial invocation to @aeval@ is passed with an indentify
function as halting continuation.
\begin{lstlisting}
def analyze(e: Expr) =
  aeval(e, Map[String, BAddr](), Store[BAddr, Storable](Map()), List(), (ans => ans))
\end{lstlisting}

@@

\begin{lstlisting}
def aeval(e: Expr, env: Env, store: BStore, time: Time, cache: Cache, cont: Cont): Ans = {
  val config = Config(e, env, store, time)
  if (cache.outContains(config)) return cont(Ans(cache.outGet(config), cache))

@@

  e match {
    case Let(x, App(f, ae), e) =>
      val closures = atomicEval(f, env, store).asInstanceOf[Set[Clos]]
      nd[Clos, Ans](closures, Ans(Set[VS](), new_cache), { case (clos, Ans(acc, cache), closnd) =>
        val Clos(Lam(v, body), c_env) = clos
        val baddr = allocBind(v, new_time)
        val new_env = c_env + (v -> baddr)
        val new_store = store.update(baddr, atomicEval(ae, env, store))
        aeval(body, new_env, new_store, new_time, cache, { case Ans(bodyvss, bodycache) =>
          nd[VS, Ans](bodyvss, Ans(Set[VS](), bodycache), { case (vs, Ans(acc_vss, cache), bdnd) =>
            val VS(vals, time, store) = vs
            val baddr = allocBind(x, time)
            val new_env = env + (x -> baddr)
            val new_store = store.update(baddr, vals)
            aeval(e, new_env, new_store, time, cache, { case Ans(evss, ecache) => bdnd(Ans(acc_vss ++ evss, ecache)) })
          },
          { case Ans(evss, cache) => closnd(Ans(evss ++ acc, cache)) })
        })
      },
      { case Ans(vss, cache) => cont(Ans(vss, cache.outUpdate(config, vss))) })

    case ae if isAtomic(ae) =>
      val vs = Set(VS(atomicEval(ae, env, store), new_time, store))
      cont(Ans(vs, new_cache.outUpdate(config, vs)))
  }
}
def analyze(e: Expr, env: Env = mtEnv, store: BStore = mtStore) = {
  def iter(cache: Cache): Ans = {
    val Ans(vss, new_cache) = aeval(e, env, store, mtTime, cache, { 
      val initConfig = Config(e, env, store, mtTime)
      case Ans(vss, cache) => Ans(vss, cache.outUpdate(initConfig, vss)) 
    })
    if (new_cache.out == cache.out) { Ans(vss, new_cache) }
    else { iter(Cache(new_cache.out, new_cache.out)) }
  }    
  iter(mtCache)
}
\end{lstlisting}

@@

\begin{lstlisting}
def aeval(e: Expr, env: Env, store: BStore, time: Time, cache: Cache): Ans = {
  val config = Config(e, env, store, time)
  if (cache.outContains(config)) {
    return Ans(cache.outGet(config), cache)
  }

  var new_cache = cache.outUpdate(config, cache.inGet(config))
  val new_time = (e::time).take(k)
  e match {
    case Let(x, App(f, ae), e) =>
      val closures = aeval(f, env, store).toList.asInstanceOf[List[Clos]]
      val letvs = for (Clos(Lam(v, body), c_env) <- closures) yield {
        val baddr = allocBind(v, new_time)
        val new_env = c_env + (v -> baddr)
        val new_store = store.update(baddr, aeval(ae, env, store))
        val Ans(bdv, bdcache) = aeval(body, new_env, new_store, new_time, new_cache)
        new_cache = bdcache
        val evs = for (VS(vals, time, store) <- bdv) yield {
          val baddr = allocBind(x, time)
          val new_env = env + (x -> baddr)
          val new_store = store.update(baddr, vals)
          val Ans(ev, ecache) = aeval(e, new_env, new_store, time, new_cache)
          new_cache = ecache
          ev
        }
        evs.foldLeft(Set[VS]())(_ ++ _)
      }
      val letv = letvs.foldLeft(Set[VS]())(_ ++ _)
      Ans(letv, new_cache.outUpdate(config, letv))
